---
title: "Project 2, parts 4 and 5: RNA-seq analysis"
output:
  pdf_document: default
  html_document: default
date: "2025-05-08"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Project 2 parts 4 and 5 overall: 75 points

Include any code that you used to solve the question as well as the
answers for all below questions. In parts 4 and 5 from project 2, you
will explore how gene expression differs between electric fish tissues.
You will learn how to analyze RNA-seq count data: identify
differentially expressed genes, distinguish samples, and make biological
interpretations. Additionally, you will visualize findings and
communicate them effectively. **Note: Please submit the Rmd and the knit
HTML file for this assignment**

### Dataset reminder:

In this project, you are be analyzing the count data and identifying
differentially expressed genes between skeletal muscle and electric
organ tissue. As a reminder, these datasets originated from two
different electric fish studies (PRJNA1005245 and PRJNA1005244). It may
help to consider how these papers analyzed the data: the methods for the
PRJNA1005244 dataset are
[published](https://doi.org/10.1093/molbev/msae021) and the methods for
the PRJNA1005245 dataset are written in the third chapter of a
[thesis](https://canvas.uoregon.edu/courses/266187/files/22059308?module_item_id=5380118).
These studies are both using the genus *Campylomormyrus*. **For this
assignment, you may not need to analyze all of the data to address the
hypothesis.**

### Resources:

There are several resources that you may use to complete this
assignment: 1) A primary resource: [DESeq2
workflow](https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#pre-filtering-the-dataset).
2) [DESeq2's additional
guide](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)
3) [Harvard Bioinformatics Core
training](https://hbctraining.github.io/DGE_workshop/lessons/04_DGE_DESeq2_analysis.html).

You are expected to use DESeq2 for this assignment. If you decide to use
an additional package, please list the package name, how you found it,
and why you chose to use it. We will do our best to help you navigate
other packages, but we cannot guarantee support for other packages.

## Project 2 part 4: Hypothesis testing and exploration

0.  In part 4, your goal is to investigate differences in gene
    expression between the electric organ and skeletal muscle. Write the
    null hypothesis for this goal below.

[There is no difference in gene expression between the electric organ and skeletal tissue.]

1.  Load your data into DESeq2

- Your data is located on Talapas:
  /projects/bgmp/shared/Bi623/Project2/counts.tar
- **Think carefully about your design formula and explain your chosen
  formula below**
- Note: you can follow the Bear_analysis rmd from class as a template or
  can also reference other resources online, such as this guide from
  [Duke](https://people.duke.edu/~ccc14/duke-hts-2017/Statistics/08032017/DESeq2-Notebook-introduction.html)
  to load your data.

```{r}
library(edgeR)
library(limma)
library(Glimma)
library(ggplot2)
library(RColorBrewer)
library(DESeq2)
library(apeglm)
library(tools)
library(qvalue)
library(dplyr)
library(data.table)
library(pheatmap)
library(PoiClaClu)
library(glmpca)
library(tibble)
library(ashr)
```

```{r}
#Include your commands, remember that you might not need to load all of the count files for this hypothesis

# Load metadata
metadata = read.csv("./Project2_Part4/Project2_RNAseq_metadata_campy.csv")
```

[My design formula is ~SpeciesSmaller + Tissue + Species_Smaller:Tissue. I am interested in seeing how the the gene expression varies between species while also controlling for tissue. I have also included in my design formula ("SpeciesSmaller:Tissue") an assumption that the expression effect of species is impacted by tissue since the tissue inherently varies by the individual it was sourced from. I am working in tandem with Zach on this as he is doing this exact design formula without this extra ':' statement so we can compare our results.]

2.  Filter the dataset to remove genes that are lowly expressed

- There are multiple ways to filter RNA-Seq datasets. You can either use
  cpm as a filter as we did in the Bear_analysis or use raw counts
  completed in the DESeq2 vignette from the above resources.
- Describe why you used your filtering strategy

```{r}
#Filter out all the baby fish since they don't have skeletal tissue
metadata = subset(metadata, metadata$AgeSize == "adult")

Campy = DESeqDataSetFromHTSeqCount(metadata,directory="Project2_Part4/counts_tar",design=~SpeciesSmaller+Tissue+SpeciesSmaller:Tissue)
```

```{r}
# Obtain CPMs
myCPM <- cpm(Campy)
# Which values in myCPM are greater than 0.5?
thresh <- myCPM > 0.5
# we would like to keep genes that have at least 2 TRUES in each row of thresh
keep <- rowSums(thresh) >= 2
# Subset the rows of count data to keep the more highly expressed genes
Campy <- Campy[keep,]
```

[I chose the counts per million approach because, similar to the study, we removed low count transcripts and only kept transcripts whose expression was above in cpm threshold in at least two separate samples. This is important to ensure that every highly expressed gene has a "replicate" in another sample to filter out outliers unique to specific samples. Unlike the study I didn't pass in a specific count threshold (they passed in 10), I used a cpm ratio of 0.5 instead per the Bear study example. Using a cpm ratio is a bit more flexible than passing in a hard count so you can eliminate low count transcripts relative to the size of the transcriptome.]

3.  Normalize the dataset and run differential expression analysis

- You may choose to normalize and run differential expression separately
  as done in the Bear_analysis, or in 1 command as done in the DESeq
  workflow 5.1.

```{r}
# Normalize the data
Campy = DESeq(Campy)
```

4.  Normalize the data to use for dimensionality reduction

- Use the rlog transformation method (follow along with the workflow,
  section 4.2)
- Describe why the rlog transformation is performed

```{r}
Campy_rlog = rlog(Campy, blind = FALSE)
```

[We used the rlog function to normalize the data in preparation to conduct and plot a principal component analysis. Without log normalization, the genes with the highest counts (and thus the highest expression) would be over-represented in these analyses and would mask any findings from the lesser expressed genes.]

5.  Identify distances between samples using sample distance heatmaps

- Generate a Euclidean distance heatmap and dendrogram using your rlog
  transformed data (follow workflow section 4.3)
- Create a Poisson distance heatmap with dendrogram using DESeq data
  (follow workflow section 4.3)
- Describe what each of these heatmaps are measuring
- Look at the sample groupings for each heatmap+dendrogram. Do you
  notice any potential individual outliers? Do both plots agree?

Note: You may need to install the packages “pheatmap” and “RColorBrewer”
to follow along. If R tells you to restart your session listen to it.

```{r}
#Include your commands for generating the Euclidean distance heatmap and dendrogram with the rlog data
campyDists = dist(t(assay(Campy_rlog)))

campyDistMatrix = as.matrix(campyDists)
rownames(campyDistMatrix) = paste(Campy_rlog$SpeciesSmaller, Campy_rlog$Tissue, sep = " - " )
colnames(campyDistMatrix) = NULL
colors <- colorRampPalette(rev(brewer.pal(9, "Greens")) )(255)
pheatmap(campyDistMatrix,
         clustering_distance_rows = campyDists,
         clustering_distance_cols = campyDists,
         col = colors)
```

```{r}
#Include your commands for generating the Poisson distance heatmap with dendrogram using your original count matrix

campyPois = PoissonDistance(t(counts(Campy)))

campyPoisDistMatrix = as.matrix(campyPois$dd)
rownames(campyPoisDistMatrix) = paste(Campy$SpeciesSmaller, Campy$Tissue, Campy$SRRfile, sep=" - " )
colnames(campyPoisDistMatrix) = NULL
pheatmap(campyPoisDistMatrix,
         clustering_distance_rows = campyPois$dd,
         clustering_distance_cols = campyPois$dd,
         col = colors)
```

[Both of these heatmaps are visualizing the clusters of expression differences in a matrix format. That's why there is a diagonal line through it- it represents the expression difference between identical samples, thus why it is 0. The plots both show that the EO and SM tissues are clustering together with the exception of one sample (Cco-SM) that is clustering with the EO samples in both plots. Of note, the Poisson plot displays more noise than the Euclidean plot.]

6.  Examine similarity in sample global gene expression using PCA,
    Generalized PCA, and MDS plots.

- Generate PCA, Generalized PCA, and MDS plots (following workflow
  section 4.4-4.6)
- Modify the color and shape of points to reflect Tissue and Species of
  samples
- Briefly describe each approach and their findings.

```{r}
#PCA
plotPCA(Campy_rlog, intgroup = c("Tissue", "SpeciesSmaller"))

```

```{r}
# Generalized PCA
gpca <- glmpca(counts(Campy), L=2)
gpca.dat <- gpca$factors
gpca.dat$SpeciesSmaller <- Campy$SpeciesSmaller
gpca.dat$Tissue <- Campy$Tissue
ggplot(gpca.dat, aes(x = dim1, y = dim2, color = SpeciesSmaller, shape = Tissue)) +
  geom_point(size =3) + coord_fixed() + ggtitle("Generalized PCA of Species and Tissue Type")
```

```{r}
# MDS plots
mds <- as.data.frame(colData(Campy_rlog))  %>%
         cbind(cmdscale(campyDistMatrix))
ggplot(mds, aes(x = `1`, y = `2`, color = SpeciesSmaller, shape = Tissue)) +
  geom_point(size = 3) + coord_fixed() + ggtitle("MDS with Campy data")
```

[All of the above approaches cluster very clearly between tissue types with the exception of the deviant skeletal muscle sample. However, they don't all differ between species equally. The MDS plot (which is the plot the researchers ended up putting in their paper) provides the clearest clustering between species followed by the PCA which mildly follows a similar clustering pattern and then the generalized PCA doesn't visualize any clustering between species. The regular PCA assumes your data follows a normal distribution while the generalized PCA does not assume this whereas the MDS approach specifically is designed for examining a distance matrix to compute pair-wise distances. The distance matrix we calculated for the heatmaps above is perfect for an MDS plot.]

7.  Use your dimensionality reduction plots, evaluate sample quality and
    complete quality adjustments.

- Based on your dimensionality reduction plots, do you have concerns
  about any samples? Explain.
- If so, remove them now, repeat steps of loading in data, filtering,
  normalization, and running the differential expression analysis.

[There is one skeletal muscle sample that consistently clustered with the electric organ samples in every dimensionality plot, so I would want to remove this sample from my metadata. See code below for this and recomputing all the calculations without it.]

```{r}
# Redo calculations without deviant skeletal muscle sample from SRR25630366 dataset

metadata = filter(metadata, metadata$SRRfile != "SRR25630366")

Campy = DESeqDataSetFromHTSeqCount(metadata,directory="Project2_Part4/counts_tar",design=~SpeciesSmaller+Tissue+SpeciesSmaller:Tissue)

# Obtain CPMs
myCPM <- cpm(Campy)
# Which values in myCPM are greater than 0.5?
thresh <- myCPM > 0.5
# we would like to keep genes that have at least 2 TRUES in each row of thresh
keep <- rowSums(thresh) >= 2
# Subset the rows of count data to keep the more highly expressed genes
Campy <- Campy[keep,]

# Normalize
Campy = DESeq(Campy)

# Normalize for dimension reduction
Campy_rlog = rlog(Campy, blind = FALSE)
```

8.  Identify genes that are differentially expressed between SM and EO
    tissues in EVERY species comparison.

- Generate a results table using log fold shrinkage and proper contrasts
- Refer to this [reference for guidance on extracting
  contrasts](https://bioinformatics-core-shared-training.github.io/Bulk_RNAseq_Course_June24/Bulk_RNAseq_Course_Base/additional_scripts_and_materials/DESeq2_results_cheatsheet.pdf)
- When setting contrasts, identify the baseline variables used in the
  intercept to help determine proper contrasts
- Filter DEGs to contain only candidates with a \|log2FoldChange\|\>1
  and adjusted value \< .05
- Join DEG data sets for every species, remove any DEG not found in
  every species, and calculate the average log2FC and average adjusted
  p-value across all species
- Visualize your cleaned DEG dataset with a [volcano
  plot](https://dputhier.github.io/jgb71e-polytech-bioinfo-app/practical/rna-seq_R/rnaseq_diff_Snf2.html).
  - Use the average log2FC and adjusted p-values for plotting
  - Make sure your plot indicates the meaning of negative and positive
    logfold changes

```{r}
## Generation of result tables

# CCO
CCO = data.frame(lfcShrink(dds = Campy, coef = "Tissue_SM_vs_EO"), type = "ashr") %>%
  rownames_to_column(var = "genes") %>%
  filter(padj < 0.05) %>%
  filter(abs(log2FoldChange) > 1) %>%
  rename("CcoLog2FoldChange" = "log2FoldChange", "Ccopadj" = "padj") %>%
  select(genes, CcoLog2FoldChange, Ccopadj)

# CCO x CRH 
CCOxCRH = data.frame(lfcShrink(dds = Campy, contrast = list(c("SpeciesSmallerCcoxCrh.TissueSM", "Tissue_SM_vs_EO")), type = "ashr")) %>%
  rownames_to_column(var = "genes") %>%
  filter(padj < 0.05) %>%
  filter(abs(log2FoldChange) > 1) %>%
  rename("CcoxCrhLog2FoldChange" = "log2FoldChange", "CcoxCrhpadj" = "padj") %>%
  select(genes, CcoxCrhLog2FoldChange, CcoxCrhpadj)

# CCO x CTS
CCOxCTS = data.frame(lfcShrink(dds = Campy, contrast = list(c("SpeciesSmallerCcoxCts.TissueSM", "Tissue_SM_vs_EO")), type = "ashr")) %>%
  rownames_to_column(var = "genes") %>%
  filter(padj < 0.05) %>%
  filter(abs(log2FoldChange) > 1) %>%
  rename("CcoxCtsLog2FoldChange" = "log2FoldChange", "CcoxCtspadj" = "padj") %>%
  select(genes, CcoxCtsLog2FoldChange, CcoxCtspadj)

#CRH
CRH = data.frame(lfcShrink(dds = Campy, contrast = list(c("SpeciesSmallerCrh.TissueSM", "Tissue_SM_vs_EO")), type = "ashr")) %>%
  rownames_to_column(var = "genes") %>%
  filter(padj < 0.05) %>%
  filter(abs(log2FoldChange) > 1) %>%
  rename("CrhLog2FoldChange" = "log2FoldChange", "Crhpadj" = "padj") %>%
  select(genes, CrhLog2FoldChange, Crhpadj)

#CTS
CTS = data.frame(lfcShrink(dds = Campy, contrast = list(c("SpeciesSmallerCts.TissueSM", "Tissue_SM_vs_EO")), type = "ashr")) %>%
  rownames_to_column(var = "genes") %>%
  filter(padj < 0.05) %>%
  filter(abs(log2FoldChange) > 1) %>%
  rename("CtsLog2FoldChange" = "log2FoldChange", "Ctspadj" = "padj") %>%
  select(genes, CtsLog2FoldChange, Ctspadj)
```

```{r}
## Joining of result tables datasets

DEG_joined_full = inner_join(x = CCO, y = CCOxCRH, by = "genes")
DEG_joined_full = inner_join(x = DEG_joined_full, y = CCOxCTS, by = "genes")
DEG_joined_full = inner_join(x = DEG_joined_full, y = CRH, by = "genes")
DEG_joined_full = inner_join(x = DEG_joined_full, y = CTS, by = "genes")

DEG_avg = DEG_joined_full %>%
  select(contains("Log2FoldChange")) %>%
  rowMeans()

DEG_padj = DEG_joined_full %>%
  select(contains("padj")) %>%
  rowMeans()

DEG_joined = DEG_joined_full %>%
  mutate(overall_avg = DEG_avg) %>%
  mutate(overall_padj = DEG_padj) %>%
  select(genes, overall_avg, overall_padj)
```


```{r}
alpha = 0.05

DEG_joined$log_padj = -log10(DEG_joined$overall_padj)


cols = densCols(DEG_joined$overall_avg, DEG_joined$log_padj)
DEG_joined$pch = 19
DEG_joined$pch[DEG_joined$pvalue ==0] = 6
plot(DEG_joined$overall_avg, 
     DEG_joined$log_padj, 
     col=cols, panel.first=grid(),
     main="Volcano Plot of SM Differential Expression with EO Baseline", 
     xlab="Effect size: log2(fold-change)",
     ylab="-log10(adjusted p-value)",
     pch=DEG_joined$pch, cex=0.4)
abline(v=0)
abline(v=c(-1,1), col="brown")
abline(h=-log10(alpha), col="brown")
```

9.  Visualize individual candidates

- Using the DEG dataframe from above, report out the top five candidates
  with the lowest average adjusted p-value.
- Using the DEG dataframe from above, visualize the normalized counts
  for the gene with the highest log2FC in the SM. Also, make a plot for
  the top candidate for the EO.

```{r}
lowest_padj = DEG_joined %>%
  select(genes, overall_padj) %>%
  arrange(overall_padj) %>%
  head(n = 5L)

highFC_gene = DEG_joined %>%
  select(genes, overall_avg) %>%
  arrange(overall_avg) %>%
  tail(n =1L)

Campy %>%
  plotCounts(gene = "snap_masked-ptg001143l-processed-gene-2.56-mRNA-1", intgroup = "Tissue", normalized = TRUE, col = Campy$Tissue)

```

10. Using the DESeq LRT test, evaluate whether there is an association
    between gene expression and electric discharge signal duration

- Complete a LRT test using design formulas to determine whether there
  is an interaction between duration and electric discharge signal
  - You may need to reload in the data to DESeq2 and re-process it
    depending on how you specified your design formula.
  - Follow this [resource to guide your
    approach](https://hbctraining.github.io/Intro-to-DGE/lessons/08a_DGE_LRT_results.html)
  - Use only the purebred species for this analysis (NO hybrids, i.e.,
    do not include ccoxcts or ccoxcrh in the analysis)
- Use degPatterns to sort genes into groups that correspond to EOD
  duration
  - From the output, identify groups that 1) have higher gene expression
    in the EO than SM and 2) have gene expression patterns that increase
    or decrease with EOD duration
  - Create a plot from degPlotCluster to showcase ONLY these selected
    groups (ex: a group of genes that has a higher abundance in longer
    compared to shorter duration electric signals)

```{r}
## YOUR CODE HERE
```

11. Based on the signal duration clustering, select 1 of the relevant
    clusters and complete a gene ontology analysis.

- Isolate a gene list for the cluster of interest
- Convert Campy gene names to D. rerio gene names using the orthology
  table derived from our ICA2 Genespace assignment (Danio_Campylomormyrus_1to1_orthologs_table.csv)
  - Note: degPatterns transforms the Campy gene names, make sure to
    change them back so they match with the orthology table
- Complete a gene ontology analysis on the cluster of interest using the
  zebrafish gene names
  - Note: enrichGO and enrichKEGG are options, use this
    [reference](https://yulab-smu.top/biomedical-knowledge-mining-book/021-go.html#clusterprofiler-go-ora)

```{r}
#Include commands
```

## Project 2 Part 5: Creating high quality figures for your report

Go back through your work from part 1-4 and identify the key pieces of
information and plots for writing your project 2 report. Remake your
chosen plots below, creating publication quality figures. Your project 2
report should have 4-6 figures, these can have multiple panels if you
know how to achieve that in R. Each figure should add a different piece
of information (i.e., do NOT include 3 dimensionality plots that show
the same pattern).

Guidelines:

-  Figures should have a title and axes that are easy to read

- Have an accessible color scheme

- Have scientifically relevant information highlighted (ex: color scheme
  denotes different species)

- When applicable, color schemes should be consistent across plots

1.  Figure 1

```{r}
#Make figure 
#OR delete this box and link a figure that you made outside of R
```

2.  Figure 2

```{r}
#Make figure 
#OR delete this box and link a figure that you made outside of R
```

3.  Figure 3

```{r}
#Make figure 
#OR delete this box and link a figure that you made outside of R
```

4.  Figure 4

```{r}
#Make figure 
#OR delete this box and link a figure that you made outside of R
```

5.  Figure 5

```{r}
#OPTIONAL analysis
```

5.  Figure 6

```{r}
#OPTIONAL analysis
```
